On the formation of H$\alpha$ line emission around classical T Tauri stars

### 2006-01-18 TTau_Halpha/kurosawa.tex
Revisions done after discussions with Tim.   (RK)
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@621 25464f21-6099-402c-b30a-640f1d40d38b

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[oneside,english,useAMS, usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amssymb}

@@

T~Tauri stars (TTS) are young ($<\sim 3\times10^{6}\,\mathrm{yrs}$,
\citealt{appenzeller:1989}) low-mass objects, and are the progenitors
of solar-type stars. Classical T~Tauri stars (CTTS) exhibit strong
H$\alpha$ emission, and typically have spectral types of F--K. Some
of the most active CTTS show emission in higher Balmer lines and metal
lines (e.g., \ion{Ca}{ii}~H and K). They also exhibit excess
 continuum flux in the ultraviolet (UV) and infrared (IR).
Their spectral energy distribution and polarisation data suggest the
presence of circumstellar discs, which plays an important role in
regulating dynamics of gas flows around CTTS (e.g. \citealt{camenzind:1990}).

Many observational studies (e.g., \citealt{herbig:1962}; \citealt{edwards:1994};
\citealt{kenyon:1994}; \citealt*{reipurth:1996}; \citealt{alencar:2000})
of CTTS line profiles have revealed evidence for both outward wind flows and
inward accretion flows, characterized by the blue-shifted absorption features
in H$\alpha$ profiles and the redshifted inverse P~Cygni (IPC) profiles
respectively. Typical mass-loss rates of CTTS are about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$
to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$ (e.g., \citealt{kuhi:1964};
\citealt{edwards:1987}; \citealt*{hartigan:1995}), and the mass-accretion
rates are also about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$ to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$
(e.g., \citealt{kenyon:1987}; \citealt*{bertout:1988}; \citealt{gullbring:1998}).
Recent H$\alpha$ spectro-astrometric observations by \citet*{takami:2003}
provide indirect evidence for the presence of bipolar and monopolar
outflows down to $\sim 1\,\mathrm{au}$ scale (e.g.\,CS~Cha and RU~Lup).
Similarly, ESO VLT observations using high-resolution ($R=50\,000$)
two-dimensional spectra of edge-on CTTS (HH30$^{*}$, HK~Tau~B,
and HV~Tau~C) by \citet{appenzeller:2005} show extended H$\alpha$
emission in the direction perpendicular to the obscuring circumstellar
disc,  suggesting the presence
of the bipolar outflows. On an even  larger scale, \emph{HST} observations
of HH30 \citep{burrows:1996} trace the jet to within $\lesssim30$~au of  the star. The jet has a cone shape with an opening angle of $3^{\circ}$
between 70 and 700 au \citep{koenigl:2000}. \citet{alencar:2000}
found about 80 per cent of their sample (30 CTTS) show blue-shifted
absorption components in at least one of the Balmer lines and Ca~K.

In the currently favoured model of accretion in CTTS, the
accretion disc is disrupted by the magnetosphere, which
channels the gas from the disc onto the stellar surface (e.g.,
\citealt{uchida:1985}; \citealt{koenigl:1991}; \citealt{cameron:1993};
\citealt{shu:1994}). This picture is supported
by recent measurements of strong ($\sim 10^{3}\mathrm{G}$)
magnetic fields in CTTS  (e.g., \citealt{johns-krull:1999};
\citealt{guenther:1996}xxxx;  \citealt{symington:2005b})
and by radiative transfer models which reproduce the gross characteristics of observed
profiles for some TTS (\citealt*{muzerolle:2001}). In particular the magnetospheric accretion (MA) model
explains blue-ward asymmetric emission line profiles as resulting from the partial occultation of
the flow by the  stellar photosphere, while inverse P~Cygni profiles in the MA model result from
inflowing material  at near free-fall velocities seen projected against hotspots on the stellar surface.

@@

In order to understand how the different parts of the CTTS circumstellar environment
contribute to the formation of H$\alpha$, the model space is divided
into four different regions: (1)~a central continuum source, (2)~the magnetospheric
accretion flow, (3)~the wind outflow, and (4)~the accretion
disc. Fig.~\ref{fig:config_dipolar} 
depicts the relative location of the regions in the model space.
The density is assumed to be rotationally symmetric around the $z$-axis. The innermost radius of
the magnetosphere at the equatorial plane coincides with the the inner
radius of the accretion disc. From the innermost part of the accretion
disc, the gas falls freely, moving along the magnetic field onto the
surface of the star.  In the following subsections we describe the details
of model components.
\begin{figure*}

@@

\caption{The basic model configuration. The system consists of four
components: (1)~the continuum source located at the origin $\left( o
\right)$ of the cartesian coordinates $\left(x,y,z\right)$ -- the
$y$-axis is into the paper, (2)~the magnetospherical accretion flow,
(3)~the bipolar wind outflow, and (4)~the accretion disc. The density
distribution is symmetric around the $z$-axis. The innermost radius of
the magnetosphere (at the equatorial plane) coincides with the the
inner radius of the accretion disc. From the innermost part of the
accretion disc, the gas falls freely, moving along the magnetic field
onto the surface of the star. The bipolar wind starts from just
outside the largest radius 
($R_{\mathrm{mo}}$) of the magnetosphere (dotted line). }

@@

\end{figure*}

@@

We adopt stellar parameters of a typical
classical T~Tauri star for the central continuum source, i.e. radius
($R_{*}$), mass ($M_{*}$), and effective temperature the photosphere
($T_{\mathrm{ph}}$) are $2\, R_{\sun}$, $0.5\, M_{\sun}$, and $4000\,\mathrm{K}$
respectively. The model atmosphere of \citet{kurucz:1993} with
$T_{\mathrm{ph}}=4000\,\mathrm{K}$ and $\log g_{*}=3.5$ (cgs) defines
the photospheric contribution to the continuum flux. The
parameters are summarised in Table~\ref{tab:std_parameters}.

@@

We adopt the MA flow model of \citet{hartmann:1994},
as done  by \citet{muzerolle:2001} and by \citet*{symington:2005},
in which the gas accretion on to the stellar surface from the innermost
part of the accretion disc occurs through a dipolar stellar magnetic
field. The magnetic field is assumed to be so strong that the gas
flow does not affect the underlying magnetic field itself. As shown
in Fig.~\ref{fig:config_dipolar}, the innermost radius ($R_{\mathrm{mi}}$)
of the magnetosphere at the equatorial plane ($z=0$) is assigned
to be same as the inner radius ($R_{\mathrm{di}}$) of the accretion
disc where the flow is truncated. In our models, $R_{\mathrm{mi}}$
and the outer radius ($R_{\mathrm{mo}}$) of the magnetosphere (at
the equatorial plane) are set to be $2.2\, R_{\sun}$ and $3.0\, R_{\sun}$
respectively. The former value is  the co-rotation radius
of the accretion disc, and the geometry of the magnetic field/stream
lines is fixed for all calculations. We note that this magnetospheric geometry
is identical to the {}``small/wide'' model of \citet{muzerolle:2001}.

The magnetic field and the gas stream lines are assumed to have the
following simple form: \begin{equation}
r=R_{\mathrm{m}}\sin^{2}\theta\label{eq:dipole}\end{equation}
 \citep*[see][]{ghosh:1977} where $r$, and $\theta$ are coordinates
of the field point ($p$) in Fig.~\ref{fig:config_dipolar} in spherical
coordinates, and $R_{\mathrm{m}}$ is the radial distance to the field
line at the equatorial plane ($\theta=\pi/2$). The range of
$R_{\mathrm{m}}$ is
restricted between $R_{\mathrm{mi}}$ and $R_{\mathrm{mo}}$. Using
the field geometry above and conservation of energy, the velocity
and the density of the accreting gas along the steam line are found
as in \citet{hartmann:1994}.

@@

In a similar fashion to the simple model of  \citet{appenzeller:2005}, the
following parametrisation of collimated bipolar wind (region 3 in
Fig.~\ref{fig:config_dipolar}) is adopted. The wind velocity field,
$\boldsymbol{v}_{\mathrm{wind}}$, consists of
radial and azimuthal components which depend on the spherical coordinates
$r$ and $\theta$. The radial component $v_{r}\left(r\right)$ is
assumed to follow the classical beta-velocity law \citep[c.f.][]{castor:1979},
and the azimuthal component $v_{\phi}$ is assumed to be a constant
fraction ($\gamma$) of the Keplerian velocity for a given distance
($w=\sqrt{x^{2}+y^{2}}$) from the symmetry axis ($z$-axis) , i.e.:

\begin{eqnarray}
\boldsymbol{v}_{\mathrm{wind}} & = & v_{r}\hat{\boldsymbol{r}}+v_{\phi}\hat{\boldsymbol{\bphi}}\label{eq:wind_velocity_def}\end{eqnarray}
where \begin{equation}
v_{r}\left(r\right)=v_{r0}+\left(v_{\infty}-v_{r0}\right)\left(1-\frac{R_{\mathrm{mo}}}{r}\right)^{\beta}\,,\label{eq:wind_velocity_radial_part}\end{equation}
\begin{equation}
v_{\phi}\left(w\right)=\gamma\,\left(\frac{GM_{*}}{w}\right)^{1/2}\,,\label{eq:wind_velocity_azimuth_part}\end{equation}
Note that the base of the wind starts at $r=R_{\mathrm{mo}}$. This naturally leads
to a density discontinuity, but allows a simple parameterization of the wind whereas
a self-consistent interface between the magnetosphere and wind would require
a full MHD calculation that is beyond the scope of this paper. The range of polar angle for
the wind is restricted to $\theta>\left|\theta_{\mathrm{disc}}\right|$
where $\theta_{\mathrm{disc}}$ is the opening angle of the accretion
disc. $v_{r0}$ is the small radial velocity
at the base of the wind ($r=R_{\mathrm{mo}}$). Normally, $v_{r0}=10\,\mathrm{km\, s^{-1}}$
which is approximately equal to the thermal velocity of hydrogen with
$T=7500\,\mathrm{K}$. Following \citet{appenzeller:2005}, $\gamma=0.05$
is adopted. The dependency of $v_{r}$ in polar direction ($\theta=0$)
on the values of wind acceleration parameter $\beta$ is shown in
Figure~\ref{fig:wind_vr_rho}. All other parameters describing the
wind are fixed as the standard values given in Table~\ref{tab:std_parameters}.

The density of the wind is assumed to be axi-symmetric and
separable in $r$ and $\theta$,

\begin{equation}
\rho\left(r,\theta\right)=P\left(r\right)\, F\left(\theta\right)\label{eq:wind_density_def}\end{equation}
 with \begin{equation}
F\left(\theta\right)=n\,\cos^{b}\theta\label{eq:wind_density_theta}\end{equation}
 where $b$ is normally a positive even number (for density symmetric
about the equatorial plane), and $n$ is the angular normalisation
constant. For $b=0$, the wind is spherically symmetric except for
the parts disrupted by the accretion disc. The larger the value of
$b$, the higher the degree of the collimation. By integrating equation~\ref{eq:wind_density_theta}
over angles and normalising the integral to $4\pi$, one finds \begin{equation}
n=\frac{1+b}{1-\cos^{1+b}\theta_{\mathrm{wind}}}\,.\label{eq:wind_density_theta_norm}\end{equation}
 Assuming the total mass-loss rate via the wind is $\dot{M}_{\mathrm{wind}}$
and the mass-flux is constant, the radial part of the density
function is reduced to $P\left(r\right)=\dot{M}_{\mathrm{wind}}\,\left[4\pi r^{2}v_{r}\left(r\right)\right]^{-1}$;
hence, equation~\ref{eq:wind_density_def} becomes \begin{equation}
\rho\left(r,\theta\right)=\frac{n\,\dot{M}_{\mathrm{wind}}\,\cos^{b}\theta}{4\pi r^{2}\, v_{r}\left(r\right)}\,.\label{eq:wind_density_final}\end{equation}
 For a given mass-accretion rate, the wind mass-loss rate in our typical
model is assigned from the ratio of mass-loss to mass-accretion rate
$\dot{M}_{\mathrm{wind}}/\dot{M}_{\mathrm{acc}}\approx0.1$, a value indicated
by both observations and MHD calculations (e.g. \citealt{koenigl:2000}).
Figure~\ref{fig:wind_vr_rho} shows the density along a streamline
in the polar direction ($\theta=0$), for different values of $\beta$
with all other parameters fixed at the standard values (Table~\ref{tab:std_parameters}).
The density is relatively sensitive to the value of $\beta$ for $r<10\, R_{\mathrm{mo}}$.

\begin{figure*}

@@

\caption{The dependency of the bipolar wind velocity and density
structure on the wind acceleration parameter $\beta$. The radial
component of the wind velocity
(equation~\ref{eq:wind_velocity_radial_part}) in polar direction,
$\theta = 0$ as a function of radius is shown on the left. The wind
density (equation~\ref{eq:wind_density_final} along polar direction as
a function of radius is shown on the right. For all $\beta$ values, the
density is normalised to the density at the base of the wind in polar
direction, i.e. $\rho_0 = \rho\left(R_\mathrm{mo},0\right)$.  The
smaller the value of $\beta$, the faster the acceleration of the
wind. For the radii $r/R_{\mathrm{mo}} < 10$, both velocity and
density are sensitive to the value of $\beta$. For all $\beta$ values,
the radial velocity of the wind reaches the terminal velocity ($\sim
210\,\kmps$) by $r/R_{\mathrm{mo}} \sim 1000$. The initial velocity
$V_{0}=10\,\kmps$, which approximately corresponds to the thermal
velocity of a hydrogen atom at 7500~K, is used for all $v_{r}$
plots.}

@@

\begin{figure}

@@

\caption{Basic model configuration of the disc-wind, magnetosphere
hybrid model. The system consist of four components: (1)~the continuum
source located at the origin $\left( o \right)$ of the cartesian
coordinates $\left(x,y,z\right)$ -- the $y$-axis is into the paper,
(2)~the magnetospherical accretion flow, (3)~the disc wind, and
(4)~the geometrically thin (but opaque) accretion disc. The
disc wind originates from the disc surface between $w_{i}=R_{wi}$ and
$w_{i}=R_{wo}$.  The wind source points ($S$), from which the stream
lines diverges, are placed at distance $d$ above and below the
star. The degree of wind collimation is controlled by changing the value of $d$. }

@@

\citet*{knigge:1995} introduced the {}``split-monopole'' kinematic
disc-wind model in their studies of the UV resonance lines formed
in the winds of cataclysmic variable stars. His formalism provides
a simple parameterization of a disc wind that has similar properties 
to those found by  MHD modelling. In this model, the outflow
arises from the surface of the rotating accretion disc, and has a
biconical geometry. The specific angular momentum is assumed to be
conserved along a stream line, and the poloidal velocity component
is assumed to be simply a radial from vertically displaced {}``sources''
from the central star.  Here we briefly describe the disc-wind model, and
readers are referred to \citet{knigge:1995} and \citet{long:2002} for
details.

The four basic parameters of the model are: (1) the mass-loss rate, (2)~the
degree of the wind collimation, (3)~the velocity gradient, and (4)~the
wind temperature. The basic configuration of the disc-wind model is
shown in Figure~\ref{fig:config_discwind}. The disc wind originates
from the disc surface, but the {}``source'' point ($S$), from which
the stream lines diverges, are placed at distance $d$ above and below
the centre of the star. The degree
of the wind collimation is controlled by changing the value of $d$. (xxxx rephrase). 
The mass-loss launching occurs between
$R_{\mathrm{wi}}$ and $R_{\mathrm{wo}}$ where the former is set
to be equal to the outer radius of the magnetosphere ($R_{\mathrm{mo}}$)
and the latter is set to $1$~au as in \citealt{krasnopolsky:2003}.

The local mass-loss rate per unit area ($\dot{m}$) is assumed to
be proportional to the mid-plane temperature of the disc, and is a
function of the cylindrical radius $w=\left(x^{2}+y^{2}\right)^{1/2}$
, i.e. \begin{equation}
\dot{m}\left(w\right)\propto T\left(w\right)^{\alpha}\,\,.\label{eq:discwind-massloss-temp}\end{equation}
 The mid-plane temperature of the disc is assumed to be expressed
as a power-law in $w$; thus, $T\propto w^{q}$. Using this in the
relation above, one finds \begin{equation}
\dot{m}\left(w\right)\propto w^{p}\label{eq:discwind-massloss-w}\end{equation}
where $p=\alpha\times q$. The index of the mid-plane temperature
power law is adopted from the dust radiative transfer model of \citet{whitney:2003a}
who found the innermost part of the accretion disc has $q=-1.15$.
In order to be consistent with the collimated disc-wind model of \citet{krasnopolsky:2003}
who used $p=-3/2$, the value of $\alpha$ is set to 1.3. The constant
of proportionality in equation~\ref{eq:discwind-massloss-w} is found
by integrating $\dot{m}$ from $R_{\mathrm{wi}}$ to $R_{\mathrm{wo}}$,
and the normalising the value to the total mass-loss rate $\dot{M}_{\mathrm{dw}}$.

The azimuthal/rotational component of the wind velocity $v_{\phi}\left(w,z\right)$
is computed from the Keplerian rotational velocity at the emerging
point of the stream line i.e. $v_{\phi}\left(w_{i},0\right)=\left(GM_{*}/w_{i}\right)^{1/2}$
where $w_{i}$ is the distance from the rotational axis ($z$) to
the emerging point on the disc, and by assuming the conservation of
the specific angular momentum along a stream line: \begin{equation}
v_{\phi}\left(w,z\right)\,=v_{\phi}\left(w_{i},0\right)\,\left(\frac{w_{i}}{w}\right)\,\,.\label{eq:discwind-toroidal-velocity}\end{equation}
Based on the canonical $\beta$ velocity law of hot stellar winds (c.f.~\citealt*{castor:1975}),
the poloidal component of the wind velocity ($v_{p}$) parameterised
as:

>>>>

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,useAMS, usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amssymb}

@@

T~Tauri stars (TTS) are young ($\gtrsim 3\times10^{6}\,\mathrm{yrs}$,
\citealt{appenzeller:1989}) low-mass objects, and are the progenitors
of solar-type stars. Classical T~Tauri stars (CTTS) exhibit strong
H$\alpha$ emission, and typically have spectral types of F--K. Some of
the most active CTTS show emission in higher Balmer lines and metal
lines (e.g., \ion{Ca}{ii}~H and K). They also exhibit excess continuum
flux in the ultraviolet (UV) and infrared (IR).  Their spectral energy
distribution and polarisation data suggest the presence of
circumstellar discs, which plays an important role in regulating
dynamics of gas flows around CTTS (e.g. \citealt{camenzind:1990}).

Many observational studies (e.g., \citealt{herbig:1962};
\citealt{edwards:1994}; \citealt{kenyon:1994};
\citealt*{reipurth:1996}; \citealt{alencar:2000}) of CTTS line
profiles have revealed evidence for both outward wind flows and inward
accretion flows, characterised by the blue-shifted absorption features
in H$\alpha$ profiles and the redshifted inverse P~Cygni (IPC)
profiles respectively. Typical mass-loss rates of CTTS are about
$10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$ to $10^{-7}\,\mathrm{M_{\sun}\,
yr^{-1}}$ (e.g., \citealt{kuhi:1964}; \citealt{edwards:1987};
\citealt*{hartigan:1995}), and the mass-accretion rates are also about
$10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$ to $10^{-7}\,\mathrm{M_{\sun}\,
yr^{-1}}$ (e.g., \citealt{kenyon:1987}; \citealt*{bertout:1988};
\citealt{gullbring:1998}).  Recent H$\alpha$ spectro-astrometric
observations by \citet*{takami:2003} provide indirect evidence for the
presence of bipolar and monopolar outflows down to $\sim
1\,\mathrm{au}$ scale (e.g.\,CS~Cha and RU~Lup).  Similarly, ESO VLT
observations using high-resolution ($R=50\,000$) two-dimensional
spectra of edge-on CTTS (HH30$^{*}$, HK~Tau~B, and HV~Tau~C) by
\citet{appenzeller:2005} show extended H$\alpha$ emission in the
direction perpendicular to the obscuring circumstellar disc,
suggesting the presence of the bipolar outflows. On an even larger
scale, \emph{HST} observations of HH30 \citep{burrows:1996} trace the
jet to within $\lesssim30$~au of the star. The jet has a cone shape
with an opening angle of $3^{\circ}$ between 70 and 700 au
\citep{koenigl:2000}. \citet{alencar:2000} found about 80 per cent of
their sample (30 CTTS) show blue-shifted absorption components in at
least one of the Balmer lines and Ca~K.

In the currently favoured model of accretion in CTTS, the accretion
disc is disrupted by the magnetosphere, which channels the gas from
the disc onto the stellar surface (e.g., \citealt{uchida:1985};
\citealt{koenigl:1991}; \citealt{cameron:1993};
\citealt{shu:1994}). This picture is supported by recent measurements
of strong ($\sim 10^{3}\mathrm{G}$) magnetic fields in CTTS (e.g.,
\citealt{johns-krull:1999}; \citealt{symington:2005b}) and by
radiative transfer models which reproduce the gross characteristics of
observed profiles for some TTS (\citealt*{muzerolle:2001}). In
particular the magnetospheric accretion (MA) model explains blue-ward
asymmetric emission line profiles as resulting from the partial
occultation of the flow by the stellar photosphere, while inverse
P~Cygni profiles in the MA model result from inflowing material at
near free-fall velocities seen projected against hotspots on the
stellar surface.

@@

In order to understand how the different parts of the CTTS
circumstellar environment contribute to the formation of H$\alpha$,
the model space is divided into four different regions: (1)~a central
continuum source, (2)~the magnetospheric accretion flow, (3)~the wind
outflow, and (4)~the accretion disc. Fig.~\ref{fig:config_dipolar}
depicts the relative location of the regions in the model space.  The
density is assumed to be rotationally symmetric around the
$z$-axis. The innermost radius of the magnetosphere at the equatorial
plane coincides with the the inner radius of the accretion disc. From
the innermost part of the accretion disc, the gas falls freely, moving
along the magnetic field onto the surface of the star.  In the
following subsections we describe the details of model components.

@@

  \caption{The basic model configuration. The system consists of four
    components: (1)~the continuum source located at the origin $\left( o
    \right)$ of the cartesian coordinates $\left(x,y,z\right)$ -- the
    $y$-axis is into the paper, (2)~the magnetospheric accretion flow,
    (3)~the bipolar wind outflow, and (4)~the accretion disc. The density
    distribution is symmetric around the $z$-axis. The innermost radius of
    the magnetosphere (at the equatorial plane) coincides with the the
    inner radius of the accretion disc. From the innermost part of the
    accretion disc, the gas falls freely, moving along the magnetic field
    onto the surface of the star. The bipolar wind starts from just
    outside the largest radius 
    ($R_{\mathrm{mo}}$) of the magnetosphere (dotted line). }

@@

@@

We adopt stellar parameters of a typical
classical T~Tauri star for the central continuum source, i.e. radius
($R_{*}$), mass ($M_{*}$), and effective temperature the photosphere
($T_{\mathrm{ph}}$) are $2\, R_{\sun}$, $0.5\, M_{\sun}$, and $4000\,\mathrm{K}$
respectively. The model atmosphere of \citet{kurucz:1979} with
$T_{\mathrm{ph}}=4000\,\mathrm{K}$ and $\log g_{*}=3.5$ (cgs) defines
the photospheric contribution to the continuum flux. The
parameters are summarised in Table~\ref{tab:std_parameters}.

@@

We adopt the MA flow model of \citet{hartmann:1994},
as done  by \citet{muzerolle:2001} and by \citet*{symington:2005},
in which the gas accretion on to the stellar surface from the innermost
part of the accretion disc occurs through a dipolar stellar magnetic
field. The magnetic field is assumed to be so strong that the gas
flow does not affect the underlying magnetic field itself. As shown
in Fig.~\ref{fig:config_dipolar}, the innermost radius ($R_{\mathrm{mi}}$)
of the magnetosphere at the equatorial plane ($z=0$) is assigned
to be same as the inner radius ($R_{\mathrm{di}}$) of the accretion
disc where the flow is truncated. In our models, $R_{\mathrm{mi}}$
and the outer radius ($R_{\mathrm{mo}}$) of the magnetosphere (at
the equatorial plane) are set to be $2.2\, R_{\sun}$ and $3.0\, R_{\sun}$
respectively. The geometry of the magnetic field/stream
lines is fixed for all calculations. We note that this magnetospheric geometry
is identical to the {}``small/wide'' model of \citet{muzerolle:2001}.

The magnetic field and the gas stream lines are assumed to have the
following simple form:
\begin{equation}
      r=R_{\mathrm{m}}\sin^{2}\theta
  \label{eq:dipole}
\end{equation}
\citep*[see][]{ghosh:1977} where $r$, and $\theta$ are coordinates
of the field point ($p$) in Fig.~\ref{fig:config_dipolar} in spherical
coordinates, and $R_{\mathrm{m}}$ is the radial distance to the field
line at the equatorial plane ($\theta=\pi/2$). The range of
$R_{\mathrm{m}}$ is restricted between $R_{\mathrm{mi}}$ and
$R_{\mathrm{mo}}$. Using the field geometry above and conservation of
energy, the velocity and the density of the accreting gas along the
steam line are found as in \citet{hartmann:1994}.

@@

In a similar fashion to the simple model of \citet{appenzeller:2005},
the following parametrisation of collimated bipolar wind (region 3 in
Fig.~\ref{fig:config_dipolar}) is adopted. The wind velocity field,
$\boldsymbol{v}_{\mathrm{wind}}$, consists of radial and azimuthal
components which depend on the spherical coordinates $r$ and
$\theta$. The radial component $v_{r}\left(r\right)$ is assumed to
follow the classical beta-velocity law \citep[c.f.][]{castor:1979},
and the azimuthal component $v_{\phi}$ is assumed to be a constant
fraction ($\gamma$) of the Keplerian velocity for a given distance
($w=\sqrt{x^{2}+y^{2}}$) from the symmetry axis ($z$-axis) , i.e.:
\begin{equation}
     \boldsymbol{v}_{\mathrm{wind}}  = 
     v_{r}\hat{\boldsymbol{r}}+v_{\phi}\hat{\boldsymbol{\bphi}}
  \label{eq:wind_velocity_def}
\end{equation}
where 
\begin{equation}
   v_{r}\left(r\right) =
   v_{r0}+\left(v_{\infty}-v_{r0}\right)\left(1-\frac{R_{\mathrm{mo}}}{r}\right)^{\beta}\,,
   \label{eq:wind_velocity_radial_part}
\end{equation}
and
\begin{equation}
      v_{\phi}\left(w\right)=\gamma\,\left(\frac{GM_{*}}{w}\right)^{1/2}\,.
   \label{eq:wind_velocity_azimuth_part}
\end{equation}
Note that the base of the wind starts at $r=R_{\mathrm{mo}}$. This
naturally leads to a density discontinuity, but allows a simple
parameterisation of the wind whereas a self-consistent interface
between the magnetosphere and wind would require a full MHD
calculation that is beyond the scope of this paper. The range of polar
angle for the wind is restricted to
$\theta>\left|\theta_{\mathrm{disc}}\right|$ where
$\theta_{\mathrm{disc}}$ is the opening angle of the accretion
disc. $v_{r0}$ is the small radial velocity at the base of the wind
($r=R_{\mathrm{mo}}$). Normally, $v_{r0}=10\,\mathrm{km\, s^{-1}}$
which is approximately equal to the thermal velocity of hydrogen with
$T=7500\,\mathrm{K}$. Following \citet{appenzeller:2005},
$\gamma=0.05$ is adopted. The dependency of $v_{r}$ in polar direction
($\theta=0$) on the values of wind acceleration parameter $\beta$ is
shown in Figure~\ref{fig:wind_vr_rho}. All other parameters describing
the wind are fixed as the standard values given in
Table~\ref{tab:std_parameters}.

The density of the wind is assumed to be axi-symmetric and
separable in $r$ and $\theta$, 
\begin{equation}
    \rho\left(r,\theta\right)=P\left(r\right)\, F\left(\theta\right)
    \label{eq:wind_density_def}
\end{equation}
with 
\begin{equation}
    F\left(\theta\right)=n\,\cos^{b}\theta
    \label{eq:wind_density_theta}
\end{equation}
where $b$ is normally a positive even number (for density symmetric
about the equatorial plane), and $n$ is the angular normalisation
constant. For $b=0$, the wind is spherically symmetric except for
the parts disrupted by the accretion disc. The larger the value of
$b$, the higher the degree of the collimation. By integrating equation~\ref{eq:wind_density_theta}
over angles and normalising the integral to $4\pi$, one finds 
\begin{equation}
  n=\frac{1+b}{1-\cos^{1+b}\theta_{\mathrm{wind}}}\,.
  \label{eq:wind_density_theta_norm}
\end{equation}
Assuming the total mass-loss rate via the wind is
$\dot{M}_{\mathrm{wind}}$ and the mass-flux is constant, the radial
part of the density function is reduced to
$P\left(r\right)=\dot{M}_{\mathrm{wind}}\,\left[4\pi
r^{2}v_{r}\left(r\right)\right]^{-1}$;
hence, equation~\ref{eq:wind_density_def} becomes 
\begin{equation}
  \rho\left(r,\theta\right) = 
  \frac{n\,\dot{M}_{\mathrm{wind}}\,\cos^{b}\theta}{4\pi r^{2}\,
    v_{r}\left(r\right)}\,.
  \label{eq:wind_density_final}
\end{equation}
For a given mass-accretion rate, the wind mass-loss rate in our
typical model is assigned from the ratio of mass-loss to
mass-accretion rate
$\dot{M}_{\mathrm{wind}}/\dot{M}_{\mathrm{acc}}\approx0.1$, a value
indicated by both observations and MHD calculations
(e.g. \citealt{koenigl:2000}).  Figure~\ref{fig:wind_vr_rho} shows the
density along a streamline in the polar direction ($\theta=0$), for
different values of $\beta$ with all other parameters fixed at the
standard values (Table~\ref{tab:std_parameters}).  The density is
relatively sensitive to the value of $\beta$ for $r<10\,
R_{\mathrm{mo}}$.

@@

  \end{center}
  \caption{The dependency of the bipolar wind velocity and density
    structure on the wind acceleration parameter $\beta$. The radial
    component of the wind velocity
    (equation~\ref{eq:wind_velocity_radial_part}) in polar direction,
    $\theta = 0$ as a function of radius is shown on the left. The wind
    density (equation~\ref{eq:wind_density_final} along polar direction as
    a function of radius is shown on the right. For all $\beta$ values, the
    density is normalised to the density at the base of the wind in polar
    direction, i.e. $\rho_0 = \rho\left(R_\mathrm{mo},0\right)$.  The
    smaller the value of $\beta$, the faster the acceleration of the
    wind. For the radii $r/R_{\mathrm{mo}} < 10$, both velocity and
    density are sensitive to the value of $\beta$. For all $\beta$ values,
    the radial velocity of the wind reaches the terminal velocity ($\sim
    210\,\kmps$) by $r/R_{\mathrm{mo}} \sim 1000$. The initial velocity
    $V_{0}=10\,\kmps$, which approximately corresponds to the thermal
    velocity of a hydrogen atom at 7500~K, is used for all $v_{r}$
    plots.}

@@

@@

  \caption{Basic model configuration of the disc-wind, magnetosphere
    hybrid model. The system consist of four components: (1)~the continuum
    source located at the origin $\left( o \right)$ of the cartesian
    coordinates $\left(x,y,z\right)$ -- the $y$-axis is into the paper,
    (2)~the magnetospheric accretion flow, (3)~the disc wind, and
    (4)~the geometrically thin (but opaque) accretion disc. The disc wind
    originates from the disc surface between $w_{i}=R_{wi}$ and
    $w_{i}=R_{wo}$.  The wind source points ($S$), from which the stream
    lines diverges, are placed at distance $d$ above and below the
    star. The degree of wind collimation is controlled by changing the
    value of $d$. }

@@

\citet*{knigge:1995} introduced the {}``split-monopole'' kinematic
disc-wind model in their studies of the UV resonance lines formed
in the winds of cataclysmic variable stars. His formalism provides
a simple parameterisation of a disc wind that has similar properties 
to those found by  MHD modelling. In this model, the outflow
arises from the surface of the rotating accretion disc, and has a
biconical geometry. The specific angular momentum is assumed to be
conserved along a stream line, and the poloidal velocity component
is assumed to be simply a radial from vertically displaced {}``sources''
from the central star.  Here we briefly describe the disc-wind model, and
readers are referred to \citet{knigge:1995} and \citet{long:2002} for
details.

The four basic parameters of the model are: (1)~the mass-loss rate,
(2)~the degree of the wind collimation, (3)~the velocity gradient, and
(4)~the wind temperature. The basic configuration of the disc-wind
model is shown in Figure~\ref{fig:config_discwind}. The disc wind
originates from the disc surface, but the {}``source'' point ($S$),
from which the stream lines diverges, are placed at distance $d$ above
and below the centre of the star.  The angle of the mass-loss
launching from the disc is controlled by changing the value of
$d$. The mass-loss launching occurs between $R_{\mathrm{wi}}$ and
$R_{\mathrm{wo}}$ where the former is set to be equal to the outer
radius of the magnetosphere ($R_{\mathrm{mo}}$) and the latter is set
to $1$~au as in \citealt{krasnopolsky:2003}.

The local mass-loss rate per unit area ($\dot{m}$) is assumed to
be proportional to the mid-plane temperature of the disc, and is a
function of the cylindrical radius $w=\left(x^{2}+y^{2}\right)^{1/2}$, i.e.
\begin{equation}
  \dot{m}\left(w\right) \propto
       T\left(w\right)^{\alpha}\,\,.\label{eq:discwind-massloss-temp}
\end{equation}
The mid-plane temperature of the disc is assumed to be expressed
as a power-law in $w$; thus, $T\propto w^{q}$. Using this in the
relation above, one finds 
\begin{equation}
  \dot{m}\left(w\right)\propto w^{p}
  \label{eq:discwind-massloss-w}
\end{equation}
where $p=\alpha\times q$. The index of the mid-plane temperature power
law is adopted from the dust radiative transfer model of
\citet{whitney:2003a} who found the innermost part of the accretion
disc has $q=-1.15$.  In order to be consistent with the collimated
disc-wind model of \citet{krasnopolsky:2003} who used $p=-3/2$, the
value of $\alpha$ is set to 1.3. The constant of proportionality in
equation~\ref{eq:discwind-massloss-w} is found by integrating
$\dot{m}$ from $R_{\mathrm{wi}}$ to $R_{\mathrm{wo}}$, and the
normalising the value to the total mass-loss rate
$\dot{M}_{\mathrm{dw}}$.
